Downstream analyses of whole-genome bisulfite-sequencing data

### 2015-04-20 Rmarkdown/datasets.Rmd
Fix formatting

\begin{table}[!htb]
\centering
\begin{tabulary}{\textwidth}{LLLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
ADS                 & Adipose                 & PE         & $75$ bp     & $23\times$    & \citet{Lister:2011kg}       \\
ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp     & $24\times$    & \citet{Lister:2011kg}       \\
ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp     & $26\times$    & \citet{Lister:2011kg}       \\
FF                  & Foreskin fibroblasts    & SE         & $85$ bp     & $16\times$    & \citet{Lister:2011kg}       \\
FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp     & $10\times$    & \citet{Lister:2011kg}       \\
FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp     & $8\times$     & \citet{Lister:2011kg}       \\
FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp     & $17\times$    & \citet{Lister:2011kg}       \\
IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
H1\_r1              & ESC                     & SE         & $85$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
H1\_r2              & ESC                     & SE         & $85$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
H1+BMP4             & IVD from H1             & SE         & $85$ bp     & $33\times$    & \citet{Lister:2011kg}       \\
H9                  & ESC                     & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
H9\_Laurent         & ESC                     & PE         & $75$ bp     & $8\times$     & \citet{Laurent:2010iw}      \\
HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
\end{tabulary}
\caption{Samples in Lister dataset. The reported read lengths are prior to any trimming of the reads. Abbreviations: \em{iPSC} = induced pluripotent stem cell; \em{ESC} = embryonic stem cell; \em{IVD} = \em{in vitro} differentiated from pluripotent cell line \em{PE} = single-end sequencing; \em{PE} = paired-end sequencing.}
\label{tab:Lister_dataset}
\end{table}

>>>>

\begin{table}[!htb]
\centering
\begin{tabulary}{\textwidth}{LLLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
ADS                 & Adipose                 & PE         & $75$ bp     & $23\times$    & \citet{Lister:2011kg}       \\
ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp     & $24\times$    & \citet{Lister:2011kg}       \\
ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp     & $26\times$    & \citet{Lister:2011kg}       \\
FF                  & Foreskin fibroblasts    & SE         & $85$ bp     & $16\times$    & \citet{Lister:2011kg}       \\
FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp     & $10\times$    & \citet{Lister:2011kg}       \\
FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp     & $8\times$     & \citet{Lister:2011kg}       \\
FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp     & $17\times$    & \citet{Lister:2011kg}       \\
IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
H1\_r1              & ESC                     & SE         & $85$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
H1\_r2              & ESC                     & SE         & $85$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
H1+BMP4             & IVD from H1             & SE         & $85$ bp     & $33\times$    & \citet{Lister:2011kg}       \\
H9                  & ESC                     & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
H9\_Laurent         & ESC                     & PE         & $75$ bp     & $8\times$     & \citet{Laurent:2010iw}      \\
HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
\end{tabulary}
\caption{Samples in Lister dataset. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
\label{tab:Lister_dataset}
\end{table}

### 2015-04-20 Rmarkdown/wgbs_downstream_analyses.Rmd
Fix formatting

\subsection{Using \em{a priori} regions}

@@

One such challenge is the size of a `MethPat` object in RAM. The `MethPat` class currently favours a simpler implementation at the expense of storage efficiency. The main inefficiency with the `MethPat` class is that the matrices storing the frequencies of each methylation pattern grow increasingly sparse as the size of the tuples increases.

Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the EPISCOPE whole-genome bisulfite-sequencing data with various sized m-tuples. Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample). However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
& \texttt{pryr::object\_size(x)} & \texttt{nrow} & Number of assays & Percentage of \texttt{NA} and $0$ values \\ \midrule
1-tuples                                 & $5.9$ GB                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$ GB                        & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{--all-combinations}) & $60.0$ GB                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$ GB                        & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$ GB                        & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
\end{tabulary}
\caption{Size of \texttt{MethPat} objects for the \em{EPISCOPE} data. All m-tuples are stranded. The size of the \texttt{MethPat} object is computed using the \texttt{object\_size} function in the \texttt{pryr} package (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The number of rows (\texttt{nrow}) corresponds to the number of m-tuples in the object. The number of assays is $2^{\text{m}}$, where $m$ is the size of the m-tuples. The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\end{table}

>>>>

\subsection{Using \emph{a priori} regions}

@@

One such challenge is the size of a `MethPat` object in memory. The `MethPat` class currently favours a simpler implementation at the expense of storage efficiency. The main inefficiency with the `MethPat` class is that the matrices storing the counts of each methylation pattern grow increasingly sparse as the size of the tuples increases.

Shown in Table \ref{tab:methpat_sparsity} is the size of the `MethPat` objects in memory for the _EPISCOPE_ whole-genome bisulfite-sequencing data with various sized m-tuples. Wee see that as the size of the m-tuples increases, the data become sparser --- most counts are 0 (meaning that particular methylation pattern was not observed in that particular sample) or `NA` (meaning that that m-tuple was not observed in that particular sample). However, for values of $m < 5$, and particularly for 'dense' data, such as RRBS, this is far less of an issue.

\begin{table}[h]
\centering
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
& Size (GB) & Number of rows & Number of assays & Percentage of $0$ and \texttt{NA} values \\ \midrule
1-tuples                                 & $5.9$                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$                       & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{{-}{-}all-combinations}) & $60.0$                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$                       & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$                       & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
\end{tabulary}
\caption{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data. All m-tuples are stranded. The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The `number of rows' corresponds to the number of m-tuples in the object. The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples. The final column is a measure of how sparse the data are --- a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\end{table}
